---
title: "Additional notes for Cox models"
author: "Devin Incerti"
---


```{r, load_libs, message = FALSE, warning = FALSE, echo = FALSE, results = 'hide'}
library(prodlim)
library(survival)
library(pec)
library(knitr)
library(tidymodels)
opts_chunk$set(comment = NA, tidy = FALSE, digits = 3,
               warning = TRUE, message = FALSE)
options(width = 100, digits = 3)
```

# Data

First, let's simulate some data:

```{r, simulate}
library(prodlim)
set.seed(43500)
train_dat <- SimSurv(200)
set.seed(2721)
test_dat <- SimSurv(2)
test_pred <- test_dat[, 5:6]
.times <- c(1, 5, 10)
```

# Cox PH models with left truncation or interval censoring
Left truncated survival data is fairly common in practice. `survival::coxph()` can fit such models and survival functions can be predicted using `survival::survfit.coxph()` or with `pec::predictSurvProb()`; however, in the latter case, a warning is returned.

```{r, cox_fit_lt}
library(survival)
library(pec)
cph_mod_lt <- coxph(Surv(time/2, time, status) ~ (X1 + X2)^2, data = train_dat,
                    x = TRUE)

# Survival predictions
summary(survfit(cph_mod_lt, newdata = test_pred), times = .times) 
predictSurvProb(cph_mod_lt, newdata = test_pred, times = .times)
```

`survival::coxph()` does not support interval censoring:

```{r, cox_interval, error = TRUE}
coxph(Surv(time/2, time, type = "interval2") ~ (X1 + X2)^2, data = train_dat)

```

# General "hack" for predicting survival from Cox models
Survival predictions can be made from a Cox model by running `survival::coxph()` with fixed parameters and then calling `survival::survfit.coxph()`. We illustrate using `mboost::glmboost()`:


```{r glm-boost-fit}
library(mboost)
f <- Surv(time, status) ~ X1 + X2
glmb_mod <- glmboost(f,data = train_dat, family = CoxPH())
```

`mboost` provides a `survFit()` function for predicting survival functions, but there is no explicit time argument:

```{r glm-boost-survFit}

glmb_surv1 <- survFit(glmb_mod, test_pred) %>%
  {.[c("time","surv")]} %>%
  do.call("cbind", .) %>%
  as_tibble()
glmb_surv1
```

We can get around this with the `survival::survfit()` "hack":
```{r, r survfit-hack}
coxfit <- survival::coxph(f, data = train_dat,
                          init = coef(glmb_mod)[-1],
                          control = survival::coxph.control(iter.max = 0))
summary(survfit(coxfit, newdata = test_pred), times = glmb_surv1$time[1:3]) 
```
