```{r, load_libs, message = FALSE, warning = FALSE, echo = FALSE, results = 'hide'}
library(prodlim)
library(survival)
library(flexsurv)
library(casebase)
library(rpart)
library(pec)
library(ipred)
library(randomForestSRC)
library(mboost)
library(tidymodels)
library(cli)
library(kableExtra)
library(knitr)
opts_chunk$set(comment = NA, tidy = FALSE, digits = 3,
               warning = FALSE, message = FALSE)
options(width = 100, digits = 3)
```

# Survival Methods in tidymodels

There is significant interest in creating infrastructure for modeling censored data using the tidymodels packages. This document is a summary and discussion of what needs to be done and a survey of many of the modeling approaches existing in R. 

We'd like to get feedback, edits, and/or contributions. Please add issues or a pull request in case for those. API suggestions are very welcome. 

# Missing components and open questions

What needs to be developed for tidymodels? `parsnip` already contains some simple wrappers for parametric survival models but, to really be useful, there is a fair amount of infrastructure needed. 

This can be broken down into a few parts: 

 1. model fit wrapper for `parsnip`
 
 2. new predictions _types_ 
 
 3. performance metrics for censored data.

The first item is fairly straight-forward. Since `parsnip` wraps model functions, this should be the easy part. 

There is some work for the second item. Many applications of censored data are less focused on predicting the actual event time; the focus is often on predicting the probability of the event (e.g. death) at specific times. For this, a new prediction type is needed. Instead of using `type = "prob"`, a different values could be used (e.g. `type = "hazard"` or `type = "survival"`). However, these predictions are _dynamic_ since they are made at specific time points. For this reason, `parsnip` and other packages will require a specific argument for the time of interest (I nominate a common argument name of `.time`). 

Since this is a big difference, it may be helpful to define a new `parsnip` model mode, perhaps called "risk prediction" or "censored regression" so that we can delineate the problem. For example, there is an `rpart` engine for `decision_tree()` that can be used for classification or regression. Without another mode, there would be conflicts or the need to make alternate function names (e.g. `survival_decision_tree()`, which is less attractive). 

The third item, performance metrics, will also require some minor modifications for dynamic predictions. A standard argument will be required for survival metrics that tell the function what column has the time of interest. 

What about `Surv()`? This function is a mainstay of survival analysis in S and R. It encapsulates the event time(s) and the censoring indicator. Most functions standardize on this format. 

The main issue is that, within data frames, it might be difficult to work with. For example:

```{r surv-obj}
library(survival)
library(tidymodels)

data(lung) # usually use `Surv(time, status)` for these data

surv_obj <- Surv(lung$time, lung$status)
is.vector(surv_obj)
is.matrix(surv_obj)

lung$surv <- surv_obj
head(lung)

lung_tbl <- as_tibble(lung)
lung_tbl$surv <- surv_obj
lung_tbl %>% slice(1:6)
```

Typically, `Surv` objects are not added as columns but are used when the model formula is specified. For tidymodels, there may be some added complexity since special, model-specific formulas might be needed to use workflows. This is very similar to special formulas for generalized additive models and mixed models.  

It would probably be better to leave the columns alone and use `Surv()` in the model formula. However, for recipes, the censoring column might get treated as a predictor or outcome so a special role might be helpful to make sure that the data are properly handled during model fitting. 
 
# Where to start


## Metrics

The cleanest place to start this work is to create `yardstick` metric functions since these are model-agnostic. The concordance statistic uses a static, event time prediction and would be very straight-forward to wrap the function from the survival package. 

For dynamic metrics, such as the prediction error curve or survival ROC metrics, the problem is a little more complex since an interface entry point is needed for the time points of interest. 

`metric_set()` will need to have a new `class` attribute. For example, depending on the type of prediction needed for a metric, this attribute is either `class_metric`, `prob_metric`, or `numeric_metric`. We'll need new classes for static and dynamic metrics.

## Model fitting

Once an official name for the new mode is determined, engines can be added for models like `surv_reg()`, `decision_tree()`, `rand_forest()`, and others. Some details are below for possible models. To begin, we should be relatively selective for picking new models/engines.  

## Prediction functions

These is a fair amount of complexity in the prediction modules. The type and format of return results from `predict()` methods can be very different and un-tidy. This alone raises the complexity of the problem. 

For some models, there will have to be different sub-modules for probability predictions that depend on the statistical distribution of the data. For example, `survreg` can use a few different distributions and translating the estimated parameters to these probabilities can be complicated. Some of this, and other details, is already handled by the `pec` package (as will be seen below). Also `mlr3` has implemented these too so, at the least, good test cases will be available.  

As previously mentioned, `predict(x, type = "survival")` will have a mandatory `.time` argument for dynamic predictions. 

Also, the convention in tidymodels is, when predicting, to produce a tibble with as many rows as `new_data` and standardized column names. For static predictions, `.pred_time` is reasonable and `.pred_survival` and `.pred_hazard` might be used for the two respective probabilities.  

If dynamic predictions are made, the result will be nested tibbles. For example, if there were two samples being predicted at three time points, the structure should look like: 

```{r nested-calcs, include = FALSE}
fitw <-
  flexsurvreg(formula = Surv(futime, fustat) ~ age,
              data = ovarian,
              dist = "weibull")
times <- 
  summary(fitw, ovarian[1:2,], type = "mean") %>% 
  map(~ setNames(.x, c("est", "lcl", "ucl"))) %>% 
  bind_rows() %>% 
  select(.pred_time = est) %>% 
  as_tibble()

probs <- 
  summary(fitw, ovarian[1:2,], type = "hazard", t = c(100, 200, 500)) %>% 
  map_dfr(~ select(.x, .time = time, .pred_hazard = est)) %>% 
  mutate(.row = rep(1:2, each = 3)) %>% 
  group_nest(.row, .key = ".pred") %>% 
  select(-.row)
  
pred_vals <- bind_cols(times, probs)
```
```{r nested-pred}
pred_vals

pred_vals$.pred[[1]]

tidyr::unnest(pred_vals, cols = c(.pred))
```

_References_: [_Evaluating Random Forests for Survival Analysis Using Prediction Error Curves._ (pdf)](https://www.jstatsoft.org/v50/i11/paper)

# Possible models

First, let's simulate some data:

```{r, simulate}
library(prodlim)
set.seed(43500)
train_dat <- SimSurv(200)
set.seed(2721)
test_dat <- SimSurv(2)
test_dat

test_pred <- test_dat[, 5:6]

# We will predict at these time points
.times <- c(1, 5, 10)
```

## Parametric Models

### `survival:::`[`survreg`](https://rdrr.io/cran/survival/man/survreg.html)

The model fitting function: 

```{r, survreg_fit}
library(survival)
sr_mod <- survreg(Surv(time, status) ~ (X1 + X2)^2, data = train_dat)
```

The `predict()` function can generate predictions for the event time and/or percentiles of the survival function via 

```{r, survreg_pred}
predict(sr_mod, test_pred)
predict(sr_mod, test_pred, type = 'quantile', p = c(.1, .5, .9))
```

Note that the latter is in terms of the time units and are not probabilities of survival by a specified time. Since these are parametric models, probability values can be derived indirectly. 

_References_: [_A package for survival analysis in R_ (pdf)](https://cran.r-project.org/web/packages/survival/vignettes/survival.pdf)


### `flexsurv::`[`flexsurvreg`](https://rdrr.io/cran/flexsurv/man/flexsurvreg.html)

The main parametric fitting function is 

```{r, flxs_fit}
library(flexsurv)
flxs_mod <-
  flexsurvreg(Surv(time, status) ~ (X1 + X2) ^ 2,
              data = train_dat,
              dist = "weibull")
```

Prediction is done using the `summary()` method and the results come back as a data frame with `tidy = TRUE`: 

```{r, flxs_pred}
# Event time prediction
summary(flxs_mod, test_pred, type = "mean", tidy = TRUE)

# Probabilities
summary(flxs_mod, test_pred, type = "survival", t = .times, tidy = TRUE)

# Same for type = "hazard"
```

The `flexsurvspline()` function has the same interface. 

_References_: [_`flexsurv`: A Platform for Parametric Survival Modelling in R_](https://www.jstatsoft.org/article/view/v070i08)


## `casebase::`[`fitSmoothHazard`](https://rdrr.io/cran/casebase/man/fitSmoothHazard.html)

The main fitting function is `fitSmoothHazard`:

```{r, cb_fit}
library(casebase)
cb_mod <-
  fitSmoothHazard(status ~ time + (X1 + X2) ^ 2,
                  data = train_dat)
```

Note that `time` is modeled explicitly and with flexibility:

```{r, cb_splines}
library(splines)
cb_sp_mod <-
  fitSmoothHazard(status ~ bs(time) + (X1 + X2) ^ 2,
                  data = train_dat)
```

And we can also do penalized estimation:

```{r, cb_glment}
cb_glmnet_mod <-
  fitSmoothHazard(status ~ time + (X1 + X2) ^ 2,
                  data = train_dat, family = "glmnet")
```

Predictions for the cumulative incidence or survival probabilities are obtained using `absoluteRisk`: 

```{r, cb_pred}
absoluteRisk(cb_mod, newdata = test_pred, time = .times, type = "CI")
absoluteRisk(cb_mod, newdata = test_pred, time = .times, type = "survival")
```

(Note that the argument `type` is only available in the development version.)

_References_: [_Fitting smooth-in-time prognostic risk functions via logistic regression_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=Fitting+smooth-in-time+prognostic+risk+functions+via+logistic+regression&btnG=), [_+A case-base sampling method for estimating recurrent event intensities_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=+A+case-base+sampling+method+for+estimating+recurrent+event+intensities&btnG=)

## Cox PH Models

### `survival:::`[`coxph`](https://rdrr.io/cran/survival/man/coxph.html)

```{r, cox_fit}
cph_mod <- coxph(Surv(time, status) ~ (X1 + X2)^2, data = train_dat, x = TRUE)
```

Both restricted mean survival time (RMST) and survival function probabilities at specific times can be predicted with `summary.survit()`:

```{r, cph_prob}
cph_surv <- summary(survfit(cph_mod, newdata = test_pred), times = .times) 

# Event time prediction (RMST)
cph_surv$table

# Survival probabilities
cph_surv 
```

Survival probabilities can also be predicted using the `pec` package;

```{r, cph_prob_pec}
predictSurvProb(cph_mod, newdata = test_pred, times = .times)
```

See the document ["cox_notes_extra"](https://github.com/tidymodels/planning/blob/main/survival-analysis/cox_notes_extra.md) file for more notes about this model. 

_References_: [_A package for survival analysis in R_ (pdf)](https://cran.r-project.org/web/packages/survival/vignettes/survival.pdf)


### `glmnet`

The model fitting input for the predictors is a matrix and the response data should be a `Surv()` object. We would have to make a new column for the interaction effect. 

```{r glmn-fit}
library(glmnet)
glmn_fit <- glmnet(x = as.matrix(train_dat[, c("X1", "X2")]), 
                   y = Surv(train_dat$time, train_dat$event),
                   family = "cox")
```

`glmnet` produces predictions across a variety of penalty values. The relative-risk can be obtained using `type = "response"`: 

```{r glmn-pred}
predict(glmn_fit, as.matrix(test_pred), type = "response")
```

_References_: [_Regularization Paths for Cox’s Proportional Hazards Model via Coordinate Descent_](https://doi.org/10.18637/jss.v039.i05)

## `mboost` models

The `mboost` package has a suite of functions for fitting boosted models whose base learners are either trees, GLM's, or GAM's. Each can take an argument of `family = CoxPH()` and each can make predictions on the linear predictor. 

For example, for a GLM model: 

```{r glm-boost-fit}
library(mboost)
glmb_mod <-
  glmboost(Surv(time, status) ~ X1 + X2,
             data = train_dat,
             family = CoxPH())
predict(glmb_mod, test_pred)
```

_References_: [_Survival ensembles_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=Survival+ensembles&btnG=)


## Tree-Based Models

### `rpart:::`[`rpart`](https://rdrr.io/cran/rpart/man/rpart.html)

```{r, rpart_fit}
library(rpart)
rp_mod <- rpart(Surv(time, status) ~ X1 + X2, data = train_dat)
```

The basic invocation of `predict` generates the predicted event time:

```{r, rpart_pred}
predict(rp_mod, test_pred)
```

The `pec` package can be used to fit an augmented `rpart` model and, from this, we can get the survivor probabilities:

```{r, rpart_prob}
pec_rpart_mod <- pecRpart(Surv(time, status) ~ X1 + X2, data = train_dat)
predictSurvProb(pec_rpart_mod, newdata = test_pred, times = .times)
```

_References_: [_An introduction to recursive partitioning using the RPART routines_ (pdf)](https://www.mayo.edu/research/documents/rpartminipdf/doc-10027257)

### `ipred:::`[`bagging`](https://rdrr.io/cran/ipred/man/bagging.html)

```{r, bag_fit}
library(ipred)
bag_mod <- bagging(Surv(time, status) ~ X1 + X2, data = train_dat)
```

When generating predictions, `survfit` objects is returned for each data point being predicted. To get survival functions including median survival:

```{r, bag_pred}
bag_preds <- predict(bag_mod, test_pred)
bag_preds

# Use summary.survfit() to get survival function (pec would also work)
# For the first sample
summary(bag_preds[[1]], times = .times)

# Median survival
library(purrr)
map_dbl(bag_preds, ~ quantile(.x, probs = .5)$quantile)
```

_References_: [_Bagging Survival Trees_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=Bagging+Survival+Trees&btnG=)


### `party:::`[`ctree`](https://rdrr.io/cran/party/man/ctree.html)


```{r, ctree_fit}
library(party)
ctree_mod <- ctree(Surv(time, status) ~ X1 + X2, data = train_dat)
```

The basic invocation of `predict` generates the predicted event time.

```{r, ctree_pred}
predict(ctree_mod, test_pred)
```

`pec` has a wrapper for this model. Since `party` uses S4 methods, this just puts the model inside of a wrapper S3 class:

```{r pec-ctree}
pec_ctree_mod <- pecCtree(Surv(time, status) ~ X1 + X2, data = train_dat)
all.equal(ctree_mod, pec_ctree_mod$ctree)

predictSurvProb(pec_ctree_mod, newdata = test_pred, times = .times)
```

_References_: [_Unbiased Recursive Partitioning: A Conditional Inference Framework_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=conditional+inference+trees+survival+models&btnG=)


### `party:::`[`cforest`](https://rdrr.io/cran/party/man/cforest.html)

```{r, cforest_fit}
library(party)

set.seed(342)
cforest_mod <-
  cforest(Surv(time, status) ~ X1 + X2,
          data = train_dat,
          controls = cforest_unbiased(ntree = 100, mtry = 1))
```

The basic invocation of `predict` generates the predicted event time: 

```{r, cforest_pred}
predict(cforest_mod, newdata = test_pred)
```

Another S3 wrapper is available: 

```{r pec-cforest}
set.seed(342)
pec_cforest_mod <-
  pecCforest(Surv(time, status) ~ X1 + X2,
             data = train_dat,
             controls = cforest_unbiased(ntree = 100, mtry = 1))

all.equal(cforest_mod, pec_cforest_mod$forest)

predictSurvProb(pec_cforest_mod, newdata = test_pred, times = .times)
```

_References_: [_Unbiased Recursive Partitioning: A Conditional Inference Framework_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=conditional+inference+trees+survival+models&btnG=)


### `randomForestSRC:::`[`rfsrc`](https://rdrr.io/cran/randomForestSRC/man/rfsrc.html)

```{r, rfsrc_fit}
library(randomForestSRC)
rfsrce_mod <- rfsrc(Surv(time, status) ~ X1 + X2, data = train_dat, ntree = 1000)
```

The `predict` function generates an object with classes `"rfsrc"`, `"predict"`, and `"surv"`. It is unclear what is being predicted:

```{r, rfsrc_pred}
predict(rfsrce_mod, test_pred)$predicted
```

The `survival` slot appears to be survival probabilities:

```{r, rfsrc_prob}
rfsrce_pred <- predict(rfsrce_mod, test_pred)
round(rfsrce_pred$survival, 2)
```

The columns appear to correspond to the "time of interest": 

```{r }
rfsrce_pred$time.interest
```

although these values do not bracket the observed times: 

```{r obs-times}
summary(train_dat$time)
```

_References_: [_Random survival forests_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=Random+survival+forests+for+R&btnG=)


## Further Candidate Models to Explore

- competing risks model via: the `cmprsk` package, `survival::survfit()`, `survival::coxph()`
- recurrent event model via: `survival::coxph()`, `rms::cph()`
- multi-state models via: `survival::coxph()` (note: no prediction method for multi-state models yet), the `mstate` and `msm` package, the `SemiMarkov` package, the `gamboostMSM` package
- joint frailty models via the `frailtypack` package
- Multi-task Logistic Regression (MTLR) via the `MLTR` package
- Bayesian model via `rstanarm::stan_surv()` (in `rstanarm` development version)


# Model Prediction Scorecard

In most cases, the open circle means that this type of prediction is facilitated _indirectly_. For example, using "tricks" related to `survfit()`, methods in the `pec` package, or other means. 

```{r pred-table, echo = FALSE, results = "asis"}
yep <- cli::symbol$tick
nope <- cli::symbol$cross
kinda <- cli::symbol$circle

pred_types <- 
  tribble(
    ~Package, ~Function, ~`pred event time`, ~`survival probs`, ~`linear pred`,
    "`survival`",        "`survreg`",        yep,  nope,  yep,
    "`flexsurv`",        "`flexsurvreg`",    yep,   yep,  yep,
    "`casebase`",        "`fitSmoothHazard`",    nope,   yep,  yep,
    "`survival`",        "`coxph`",         yep, yep,  yep,
    "`glmnet`",          "`glmnet`",        kinda, kinda,  yep,
    "`mboost`",          "`*boost`",        kinda, kinda,  yep,
    "`rpart`",           "`rpart`",          yep, kinda, nope,
    "`ipred`",           "`bagging`",        yep, kinda, nope,
    "`party`",           "`ctree`",         yep,  kinda, nope,
    "`party`",           "`cforest`",       yep,  kinda, nope,
    "`randomForestSRC`", "`rfsrc`",         yep,  kinda, nope
  )

pred_types %>% 
  kable(escape = FALSE) %>% 
  kableExtra::kable_styling()
```

