---
title: "OOB as a metric"
format: pdf
execute:
  warning: false
  message: false
---

# Out of Bag Errors

For bagged models in general, but for Random Forests in particular.

## Why do we care about OOB errors?

I regularly teach a suite of ML and statistical models in an applied upper division statistics course.  There are two reasons why I would like to harness the OOB errors.

### 1. OOB errors are more efficient than CV

If we can tune hyper parameters using OOB errors instead of CV, the model will be an order of magnitude more efficient.

### 2. OOB errors are part of the pedagogical process

One way to understand something is to understand it in contrast to anothe thing. Contrasting OOB and CV deepens the understanding of both metrics: how they are computed, used, and interpreted.

# Example

```{r}
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(randomForest)
```

```{r}
set.seed(47)
penguins <- penguins |>
    drop_na(body_mass_g)

penguin_split <- initial_split(penguins)
penguin_train <- training(penguin_split)
penguin_test <- testing(penguin_split)
```

```{r}
penguin_bag_recipe <-
  recipe(body_mass_g ~ . ,
         data = penguin_train) |>
  step_unknown(sex, new_level = "unknown") |>
  step_mutate(year = as.factor(year)) |>
  step_naomit(all_predictors(), body_mass_g)


num_trees <- 500
penguin_bag <- rand_forest(mtry = 7,
                           trees = num_trees) |>
  set_engine("randomForest", oob.error = TRUE) |>  # note that we keep the OOB error!
  set_mode("regression")


penguin_bag_wflow <- workflow() |>
  add_model(penguin_bag) |>
  add_recipe(penguin_bag_recipe)


penguin_bag_fit <- 
  penguin_bag_wflow |>
  fit(data = penguin_train)

```


## Plotting OOB

Using `extract_fit_parsnip()` and `pluck()` we can pull out the relevant OOB metrics (either $R^2$ or MSE), but only as a function of **number of tres** (and not other tuning parameters).

(Tbh, I'm not 100% sure that the values plotted below are OOB, but I think they are because of the `oob.error = TRUE` argument above.)

```{r}
penguin_bag_fit |>
  extract_fit_parsnip() 
```

```{r}
rsq <- penguin_bag_fit |>
  extract_fit_parsnip() |>
  pluck("fit") |>
  pluck("rsq")


data.frame(
  trees = 1:num_trees, 
  rsq = rsq) |>
  ggplot() + 
  geom_line(
    aes(x = trees, y = rsq)) +
  ylab("") + 
  xlab("Number of Trees") + 
  ggtitle("OOB R-squared")
```


```{r}
MSE <- penguin_bag_fit |>
  extract_fit_parsnip() |>
  pluck("fit") |>
  pluck("mse")

data.frame(
  trees = 1:num_trees, 
  MSE = MSE) |>
  ggplot() + 
  geom_line(
    aes(x = trees, y = MSE)) +
  ylab("") + 
  xlab("Number of Trees") + 
  ggtitle("OOB MSE")
```

## Model fitting using CV

We cannot easily, however, use OOB to do any of the model fitting. Consider the process below which selects `mtry` for a Random Forest model. The metric used is `rmse`, but the only way (as far as I can tell) to run the analysis is through cross validation. Because the OOB models have already been calculated, I would think that using CV is overkill and a waste of computational resources.


```{r}
penguin_rf_recipe <-
  recipe(body_mass_g ~ . ,
         data = penguin_train) |>
  step_unknown(sex, new_level = "unknown") |>
  step_mutate(year = as.factor(year)) 

penguin_rf <- rand_forest(mtry = tune(),
                           trees = tune()) |>
  set_engine("ranger") |>
  set_mode("regression")

penguin_rf_wflow <- workflow() |>
  add_model(penguin_rf) |>
  add_recipe(penguin_rf_recipe)

set.seed(34)
penguin_folds <- vfold_cv(penguin_train,
                          v = 4)

penguin_grid <- grid_regular(mtry(range = c(2,7)),
                             trees(range = c(1,500)), levels = 5)                          

penguin_rf_tune <- 
  penguin_rf_wflow |>
  tune_grid(resamples = penguin_folds,
            grid = penguin_grid)

```

We use the `rmse` metric to consider the best `mtry`.  Would be ideal to use the OOB rmse for such a task. In this case, it seems like `mtry` of 2 out performs the other `mtry` values, using cross validation.

```{r}
penguin_rf_tune |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  mutate(mtry = as.factor(mtry)) |> 
  ggplot() + 
  geom_line(aes(x = trees, y = mean, color = mtry)) + 
  labs(title = "RMSE: root mean squared error", y = "")
```

The point I'm trying to make is that cross validation is overkill when you have OOB information. The `rmse` based on the OOB models could be calculated when building the models, and so we wouldn't need to re-run everything 4 times for 4-fold cross validation.


## Reporting metrics

Similarly, it would be nice to just be able to "pick" and report the best model using OOB errors.  Would be ideal to have OOB rmse and OOB rsq as options here.

```{r}
finalize_workflow(
    penguin_rf_wflow,
  select_best(penguin_rf_tune, metric = "rmse")) |>
    last_fit(split = penguin_split) |>
    collect_metrics()
```

# Model fitting using OOB

@EmilHvitfeldt [wrote the function](https://github.com/tidymodels/planning/issues/25#issuecomment-1458731842) `extract_oob()` (below), and it doesn't seem to be a perfectly parallel construction to `collect_metrics()`.  Although, I think I can get the same results with some tidyverse wrangling.  

How do we know that we have the OOB error? And is it rmse? Is there an equivalent $R^2$?

```{r}
extract_oob <- function(x) {
  x |>
    extract_fit_engine() |>
    pluck("prediction.error")
}

penguin_rf_tune_oob <- 
  penguin_rf_wflow |>
  tune_grid(resamples = penguin_folds,
            grid = penguin_grid,
            control = control_grid(extract = extract_oob)) |>
                unnest(.extracts) |>
                unnest(.extracts)

penguin_rf_tune_oob 
```

I think I would parse the code (ala `collect_metrics()`) using:

```{r}
penguin_rf_tune_oob |>
  group_by(mtry, trees) |>
  summarize(mean_oob = mean(.extracts), .groups = "drop") |>
  arrange(mean_oob)
```
